
---
title: "Weather Impact Analysis Across the United States"
author: "Vincent Buzzano"
date: "Monday, November 24, 2014"
output: 
  html_document:
    keep_md: true
---

## Synopsis

In this analyse, I will try to address two questions:  

 1. Across the United States, which types of events are most harmful with respect to population health?

 2. Across the United States, which types of events have the greatest economic consequences?


## Data Processing
Here we go, we download the dataset [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb], read it and clean it.

There is also some documentation of the database available.   
 
 - [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)  
 - [National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)  


```{r echo = TRUE, cache = TRUE}

# Download dataset
if (!file.exists('StormData.csv.bz2'))
    download.file(url = 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2', destfile = 'StormData.csv.bz2', method = 'curl')

# read data
data <- read.table(bzfile("StormData.csv.bz2"), header = TRUE, sep = ',')

# keep only data with fatalities, injuries or propdmg value greater than 0
data <- data[data$FATALITIES > 0 | data$INJURIES > 0 | data$PROPDMG > 0, ]
```

Then we group all events by type by truncating their values to 3 caracters.

     EVENT TYPES                  GROUP
     TORNADO DEBRIS            -> TOR
     TORNADO F0                -> TOR        
     TORNADO F1                -> TOR

     URBAN SMALL               -> URB
     URBAN SMALL STREAM FLOOD  -> URB
     URBAN/SMALL               -> URB

```{r echo = TRUE, cache = TRUE}
# create Group
evTypes <- levels(data$EVTYPE)
data$EVTYPE_GROUP <- strtrim(gsub(" ", "", toupper(data$EVTYPE)), 3)
```


### Processing for Question 1

I get total of fatalities and injuries for each events types group, sum them and prepare dataset to plot the top 10 and report the mote harmful type.  

```{r echo = TRUE, cache = TRUE}
# Question 1
# aggregate and sum FATALITIES for each EVTYPE GROUP
fatalities <- aggregate(FATALITIES~EVTYPE_GROUP, data, sum)
# aggregate and sum INJURIES for each EVTYPE GROUP
injuries <- aggregate(INJURIES~EVTYPE_GROUP, data, sum)

totalFatalities         <- sum(data$FATALITIES)
totalInjuries           <- sum(data$INJURIES)
ratioFatalitiesInjuries <- totalInjuries / totalFatalities

t1 <- fatalities[, c(1, 2)]
t1$INJURIES <- injuries[injuries$EVTYPE_GROUP ==t1$EVTYPE_GROUP, 2]
t1$TOTAL <- t1$FATALITIES + t1$INJURIES
t1 <- t1[with(t1, order(-TOTAL)), ]

# get top 10
result1 <- head(t1$TOTAL, 10)
names(result1) <- head(t1$EVTYPE_GROUP, 10)

# get top 10
t2 <- t1
t2$TOTAL <- (t2$FATALITIES * ratioFatalitiesInjuries) + t2$INJURIES
t2 <- t2[with(t2, order(-TOTAL)), ]

gGroup = t1[1,'EVTYPE_GROUP']

result2 <- head(t2$TOTAL, 10)
names(result2) <- head(t2$EVTYPE_GROUP, 10)

# get the most harmful type for question 1
mostHarmfulType <- evTypes[grepl(paste("^", gGroup, sep=''), toupper(evTypes))][1]
```

### Processing for Question 2
I create a new field multiple to multiply the amount of damage for each line of events. Then prepare the dataset to plot the top 10 and report the mote harmful type.  

```{r echo = TRUE, cache = TRUE}
# Question 2
data$multiple <- 1
data$multiple[data$PROPDMGEXP =="K"]  <- 1000
data$multiple[data$PROPDMGEXP =="M"]  <- 1000000
data$multiple[data$PROPDMGEXP =="B"]  <- 1000000000
for(i in c(1:10))
    data$multiple[data$PROPDMGEXP == as.character(i)]  <- as.numeric(paste(1,paste(rep(0, 5), collapse = ''), sep = ''))
data$PROPDMG2 <- data$PROPDMG * data$multiple

# aggregate and sum properties damage for each EVTYPE GROUP
propdmg <- aggregate(PROPDMG2~EVTYPE_GROUP, data, sum)
propdmg <- propdmg[with(propdmg, order(-PROPDMG2)), ]

gGroup2 = propdmg[1,'EVTYPE_GROUP']

# get top 10
result3 <- head(propdmg$PROPDMG2, 10)
names(result3) <- head(propdmg$EVTYPE_GROUP, 10)

# get most harmful type for question 2
mostHarmfulType2 <- evTypes[grepl(paste("^", gGroup2, sep=''), toupper(evTypes))][1]

```

## Results

### Question 1
We can now answer to Question 1.

```{r echo = TRUE}
barplot(result1, xlab = 'Events Type', ylab='sum of fatalities and injuries', main="Top 10 of events type which is the most harmful to population health");
```


I know there is a total of **`r totalFatalities` fatalities** against a total of **`r totalInjuries` injuries**. I can try to apply the ratio *injuries*/*fatalities* **`r ratioFatalitiesInjuries`** to fatatlities to adjust our results. I think that death is worse than a corporal domage.

```{r echo = TRUE}
barplot(result2, xlab = 'Events Type', ylab='sum of fatalities and injuries', main="Top 10 of events type which is the most harmful to population health");
```

The types of events which is the most harmful to population health is: **`r mostHarmfulType`**



### Question 2

The answer of Question 2

```{r echo = TRUE}
barplot(result3, xlab = 'Events Type', ylab='sum of properties damage in $', main="Top 10 of events type that have the greatest economic consequences");
```

The types of events that have the greatest economic consequences is: **`r mostHarmfulType2`**

